The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-03-05 pkg/yuimadocs/inst/doc/JSS/article-new.Rnw
update revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@233 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{User Specified State and Time Variables: NEW SECTION}
Suppose now that the user wants to specify her own model using a prescribed notation, e.g., some sdes like
$$\de Y_s = -3 s Y_s \de s + \frac{1}{1+Y_s^2}\de W_s,$$
where $a(s,y) = -3sy$ and $b(y) = 1/(1+y^2)$.
Then this model can be described in \pkg{yuima} as follows
<<echo=TRUE, print=FALSE,results=hide>>=
mod1b <- setModel(drift = "-3*s*y", diffusion = "1/(1+y^2)", 
state.var="y", time.var="s")
@
In this case the solution variable is the same as the state variable. Indeed, the \code{yuima.model} object appears as follows
<<>>=
str(mod1b)
@
Once again, the user may use the \code{simulate} method to perform simulation.

@@

Of course, \pkg{yuima} approach is more general in that the above Levy approximation only holds when the process $X_t$ is the geometric Brownian motion. We now give an example when the underlying process $X_t$ is the following CIR model
$$
\de X_t =  0.9 X_t \de t + \ep \sqrt{X_t} \de W_t, X0 = 1
$$
and we calculate the asymptotic expansion of a european call option with strike price $K=10$ for $\ep=0.4$.

@@

time1 <- proc.time()
asymp <- asymptotic_term(yuima.ae,block=100,rho,g)
time2 <- proc.time()
@
We now extract the first and second order terms of the asymptotic expansion from the \code{asymp} object
<<>>=
ae.value0 <- asymp$d0
ae.value0
ae.value1 <- asymp$d0 + e * asymp$d1
ae.value1
ae.value2 <- as.numeric(asymp$d0 + e * asymp$d1 + e^2 * asymp$d2)
ae.value2
ae.time <- time2 - time1
ae.time
@
As it can be seen, the contribution of the term corresponding to the second order of the asymptotic expansion give a real contribution to the approximation and the final approximated value \Sexpr{round(ae.value2,5)} can be compared with a Monte Carlo experiment with 1000000 replications
<<>>=
time3 <- proc.time()

time4 <- proc.time()
mc.time <- time4 - time3
mc.time/ae.time*100
@

@@

\subsection{How to input data into a \code{yuima} object}
Although most of the examples in this section are given on simulated data, the main way to fill up the \code{data} slot of a \code{yuima} object is to use the function \code{setYuima}. The function \code{setYuima} sets various slots of the \code{yuima} object. In particular, to estimate a \code{yuima.model} called \code{mod} on the data \code{X} one can use a code like this \code{my.yuima <- setYuima(data=setData(X), model=mod)} and then pass \code{my.yuima} to the inference functions as described in what follows.

@@

\subsection{Quasi Maximum Likelihood Estimation}
Consider a multidimensional diffusion process
\begin{equation}
\label{eq:sdemle}
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t,\quad X_0=x_0,
\end{equation}
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial variable $X_0$. 
Moreover, $\theta_1\in\Theta_1\subset \mathbb{R}^p$, 
$\theta_2\in\Theta_2\subset \mathbb{R}^q$,
$a:\mathbb{R}^d\times\Theta_2 \to \mathbb{R}^d$ and $b:\mathbb{R}^d\times\Theta_1 \to \mathbb{R}^d\times \mathbb{R}^r$. 
The naming of $\theta_2$ and $\theta_1$ is theoretically natural in view of the optimal convergence rates of the estimators for these parameters.
Given sampled data ${\bf X}_n = (X_{t_i})_{i=0,\ldots,n}$ with $t_i = i\Delta_n$,
QMLE (Quasi Maximum Likelihood Estimator) makes use of the following approximation of the true log-likelihood for multidimensional diffusions
{\small
\begin{eqnarray}\label{qlik}
\ell_n({\bf X}_n,\theta)
=-\frac12\sum_{i=1}^n\left\{\log\text{det}(\Sigma_{i-1}(\theta_1))
+\frac{1}{\Delta_n}
\Sigma_{i-1}^{-1}(\theta_1)[(\Delta X_i-\Delta_n a_{i-1}(\theta_2))^{\otimes 2}]\right\}
\end{eqnarray}}
\noindent
where $\theta=(\theta_1, \theta_2)$, $\Delta X_i=X_{t_i}-X_{t_{i-1}}$, $\Sigma_i(\theta_1)=\Sigma(\theta_1,X_{t_i})$, $a_i(\theta_2)=a(X_{t_i},\theta_2)$, $\Sigma=b^{\otimes 2}$, $A^{\otimes 2}= A A^T$ and $A^{-1}$ the inverse of $A$, $A[B] = \mbox{tr}(A B)$.  
Then 
%\citep[see e.g.,][]{Yoshida92, Kessler97, Yoshida11}, 
the QMLE of $\theta$ is an estimator that satisfies  
$$\hat\theta=\arg\max_\theta \ell_n({\bf X}_n,\theta)$$
exactly or approximately.

@@

\subsubsection{QMLE in practice}

@@

\subsubsection{Theoretical remarks on QMLE}%, High-freq \& ergodicity}
Here are some theoretical remarks. 
Consistency must be required when an estimator is applied; otherwise 
the estimation would lose mathematical backing because 
the more data the observer obtains, the worse the estimator behaves. 
For consistency of $\hat{\theta}_1$, we are assuming 
$\Delta_n\to0$ as $n\to\infty$. Indeed, under this condition, $\hat{\theta}_1$ 
has asymptotically (mixed) normality 
\citep{GenonCatalotJacod93, UchidaYoshida12}. 
On the other hand, one needs $T=n\Delta_n\to\infty$ for consistency of 
$\hat{\theta}_2$ because 
the Fisher information for $\theta_2$ is finite for a finite $T$ 
and consistent estimation of $\theta_2$ is theoretically impossible. 
When $T\to\infty$, usually ergodicity is assumed to ensure a law of large numbers and as a result 
we obtain 
consistency of $\hat{\theta}_2$, moreover asymptotic normality. 
We assume the condition $n\Delta_n^p\to0$ for $p=2$ 
while applying the quasi log-likelihood (\ref{qlik}) based on the local Gaussian approximation. 
In practical applications, reduction of the parameter's dimension and relaxing 
the above condition to $n\Delta_n^p\to0$ for $p$ bigger than $2$ are very important. 
Adaptive estimation methods were proposed for $p=3$ and for any $p$ 
in \citep{Yoshida92} and \citep{UchidaYoshida11}, respectively, with the convergence of moments 
by a large deviation argument. 
When $T$ is regarded to be not large, the small sample effect on estimation of $\theta_2$ appears, which 
will be discussed in Section \ref{small.sample}. 
Modules for QMLE and Bayes estimators are going to be available for jump-diffusions. 
See Shimizu and Yoshida \citep{ShimizuYoshida2006} 
and Ogihara and Yoshida \citep{OgiharaYoshida2011}.

@@

As said, the change-point analysis do not consider the information coming from the drift part of the model and \pkg{yuima} ignores this internally.
Just to make clear that the information on the drift term is not considered by the function \code{CPoint}, we redefine the \code{yuima} model removing the information coming from the drift and then adding back the data.
<<cpoint4b>>=
noDriftModel <- setModel(drift=c("0", "0"), diffusion=diff.matrix,
 time.variable="t", state.variable=c("x1", "x2"), 
 solve.variable=c("x1", "x2"))
noDriftModel <- setYuima(noDriftModel, data=yuima@data)
noDriftModel@model@drift
@
First we show that there is no difference in using the complete model or the model without drift. For simplicity, we assume the true values of the parameters for $\theta_{1.k}$ and $\theta_{1.k}$
<<cpoint5>>=
t.est <- CPoint(yuima,param1=t0,param2=t1)
t.est$tau
t.est2 <- CPoint(noDriftModel,param1=t0,param2=t1)
t.est2$tau
@
\subsubsection{An example of two stage estimation}
In practical situations, the initial values of the parameters are not known and there is the need to provide a preliminary estimators of them. One possible approach is the two stage change-point estimation approach as explained in  \citet{iacyos09}. The idea is to take a small subsets of observations at the very beginning and the end of the time series, estimate a
change point and and then refine the estimation.

>>>>

\subsection{User Specified State and Time Variables}
Suppose now that the user wants to specify her own model using a prescribed notation, e.g., some sdes like
$$\de Y_s = -3 s Y_s \de s + \frac{1}{1+Y_s^2}\de W_s,$$
where $a(s,y) = -3sy$ and $b(y) = 1/(1+y^2)$.
Then this model can be described in \pkg{yuima} as follows
<<echo=TRUE, print=FALSE,results=hide>>=
mod1b <- setModel(drift = "-3*s*y", diffusion = "1/(1+y^2)", 
state.var="y", time.var="s")
@
In this case the solution variable is the same as the state variable. Indeed, the \code{yuima.model} object appears as follows
<<>>=
str(mod1b)
@
Once again, the user may use the \code{simulate} method to perform simulation.

@@

\subsection{Asymptotic Expansion for General Stochastic Processes}
Of course, \pkg{yuima} approach is more general in that the above Levy approximation only holds when the process $X_t$ is the geometric Brownian motion. We now give an example when the underlying process $X_t$ is the following CIR model
$$
\de X_t =  0.9 X_t \de t + \ep \sqrt{X_t} \de W_t, X0 = 1
$$
and we calculate the asymptotic expansion of a european call option with strike price $K=10$ for $\ep=0.4$.

@@

time1 <- proc.time()
asymp <- asymptotic_term(yuima.ae,block=100,rho,g)
time2 <- proc.time()
@
We now extract the first and second order terms of the asymptotic expansion from the \code{asymp} object
<<>>=
ae.value0 <- asymp$d0
ae.value0
ae.value1 <- asymp$d0 + e * asymp$d1
ae.value1
ae.value2 <- as.numeric(asymp$d0 + e * asymp$d1 + e^2 * asymp$d2)
ae.value2
ae.time <- time2 - time1
ae.time
@
As it can be seen, the contribution of the term corresponding to the second order of the asymptotic expansion give a real contribution to the approximation and the final approximated value \Sexpr{round(ae.value2,5)} can be compared with a Monte Carlo estimate based on 1000000 replications 
which is equal to \code{0.561059}, but more demanding in terms of cpu time. The relative difference among the two estimates is \Sexpr{round((ae.value2-0.561059)/0.561059*100,1)}\%.

@@

\subsection{How to Input Data Into a \code{yuima} Object}
Although most of the examples in this section are given on simulated data, the main way to fill up the \code{data} slot of a \code{yuima} object is to use the function \code{setYuima}. The function \code{setYuima} sets various slots of the \code{yuima} object. In particular, to estimate a \code{yuima.model} called \code{mod} on the data \code{X} one can use a code like this \code{my.yuima <- setYuima(data=setData(X), model=mod)} and then pass \code{my.yuima} to the inference functions as described in what follows.

@@

\subsection{Quasi Maximum Likelihood Estimation}\label{section-qmle}
Consider a multidimensional diffusion process
\begin{equation}
\label{eq:sdemle}
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t,\quad X_0=x_0,
\end{equation}
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial variable $X_0$. 
Moreover, $\theta_1\in\Theta_1\subset \mathbb{R}^p$, 
$\theta_2\in\Theta_2\subset \mathbb{R}^q$,
$a:\mathbb{R}^d\times\Theta_2 \to \mathbb{R}^d$ and $b:\mathbb{R}^d\times\Theta_1 \to \mathbb{R}^d\times \mathbb{R}^r$. 
The naming of $\theta_2$ and $\theta_1$ is theoretically natural in view of the optimal convergence rates of the estimators for these parameters.
Given sampled data ${\bf X}_n = (X_{t_i})_{i=0,\ldots,n}$ with $t_i = i\Delta_n$,
QMLE (Quasi Maximum Likelihood Estimator) makes use of the following approximation of the true log-likelihood for multidimensional diffusions
{\small
\begin{eqnarray}\label{qlik}
\ell_n({\bf X}_n,\theta)
=-\frac12\sum_{i=1}^n\left\{\log\text{det}(\Sigma_{i-1}(\theta_1))
+\frac{1}{\Delta_n}
\Sigma_{i-1}^{-1}(\theta_1)[(\Delta X_i-\Delta_n a_{i-1}(\theta_2))^{\otimes 2}]\right\}
\end{eqnarray}}
\noindent
where $\theta=(\theta_1, \theta_2)$, $\Delta X_i=X_{t_i}-X_{t_{i-1}}$, $\Sigma_i(\theta_1)=\Sigma(\theta_1,X_{t_i})$, $a_i(\theta_2)=a(X_{t_i},\theta_2)$, $\Sigma=b^{\otimes 2}$, $A^{\otimes 2}= A A^T$ and $A^{-1}$ the inverse of $A$, $A[B] = \mbox{tr}(A B)$.  
Then 
%\citep[see e.g.,][]{Yoshida92, Kessler97, Yoshida11}, 
the QMLE of $\theta$ is an estimator that satisfies  
$$\hat\theta=\arg\max_\theta \ell_n({\bf X}_n,\theta)$$
exactly or approximately.

@@

\subsubsection{QMLE in Practice}

@@

\subsubsection{Theoretical Remarks on QMLE}%, High-freq \& ergodicity}
Here are some theoretical remarks. 
Consistency must be required when an estimator is applied; otherwise 
the estimation would lose mathematical backing because 
the more data the observer obtains, the worse the estimator behaves. 
For consistency of $\hat{\theta}_1$, we are assuming 
$\Delta_n\to0$ as $n\to\infty$. Indeed, under this condition, $\hat{\theta}_1$ 
has asymptotically (mixed) normality 
\citep{GenonCatalotJacod93, UchidaYoshida12}. 
On the other hand, one needs $T=n\Delta_n\to\infty$ for consistency of 
$\hat{\theta}_2$ because 
the Fisher information for $\theta_2$ is finite for a finite $T$ 
and consistent estimation of $\theta_2$ is theoretically impossible. 
When $T\to\infty$, usually ergodicity is assumed to ensure a law of large numbers and as a result 
we obtain 
consistency of $\hat{\theta}_2$, moreover asymptotic normality. 
We assume the condition $n\Delta_n^p\to0$ for $p=2$ 
while applying the quasi log-likelihood (\ref{qlik}) based on the local Gaussian approximation. 
In practical applications, reduction of the parameter's dimension and relaxing 
the above condition to $n\Delta_n^p\to0$ for $p$ bigger than $2$ are very important. 
Adaptive estimation methods were proposed for $p=3$ and for any $p$ 
in \citep{Yoshida92} and \citep{UchidaYoshida11}, respectively, with the convergence of moments 
by a large deviation argument. 
When $T$ is regarded to be not large, the small sample effect on estimation of $\theta_2$ appears, which 
will be discussed in Section \ref{small.sample}. 
Modules for QMLE and Bayes estimators are going to be available for jump-diffusions. 
See Shimizu and Yoshida \citep{ShimizuYoshida2006} 
and Ogihara and Yoshida \citep{OgiharaYoshida2011}.

@@

As said, the change-point analysis do not consider the information coming from the drift part of the model and \pkg{yuima} ignores this internally.
Just to make clear that the information on the drift term is not considered by the function \code{CPoint}, we redefine the \code{yuima} model removing the information coming from the drift and then adding back the data.
<<cpoint4b>>=
noDriftModel <- setModel(drift=c("0", "0"), diffusion=diff.matrix,
 time.variable="t", state.variable=c("x1", "x2"), 
 solve.variable=c("x1", "x2"))
noDriftModel <- setYuima(noDriftModel, data=yuima@data)
noDriftModel@model@drift
@
First we show that there is no difference in using the complete model or the model without drift. For simplicity, we assume the true values of the parameters for $\theta_{1.k}$ and $\theta_{1.k}$
<<cpoint5>>=
t.est <- CPoint(yuima,param1=t0,param2=t1)
t.est$tau
t.est2 <- CPoint(noDriftModel,param1=t0,param2=t1)
t.est2$tau
@
\subsubsection{An Example of Two Stage Estimation}
In practical situations, the initial values of the parameters are not known and there is the need to provide a preliminary estimators of them. One possible approach is the two stage change-point estimation approach as explained in  \citet{iacyos09}. The idea is to take a small subsets of observations at the very beginning and the end of the time series, estimate a
change point and and then refine the estimation.
